BOLD Parameter Estimation Using Particle Filters

### 2010-08-15 doc/thesis/results.tex
low/high noise signal case reviewed

To test how well the particle filter would do plenty of data, and 
determine the inherent variance in the model parameters, a very long simulation
with randomly generated impulse stimulus was created. The preprocessed timeseries
and the final estimate are shown in \autoref{fig:long_converge};
the final covariance matrix of the parameters is in \autoref{tab:long_cov}.
Note that even though the BOLD response converged well (\autoref{fig:long_converge}),
the parameters still have significant correlation (\autoref{tab:long_corr}). 
This is in spite of the noisy input (green line in \autoref{fig:long_converge}).
Based on the histograms in the first 500 seconds, the parameters converged to
their final values well before the end.
Although this is only a single test, the correlation (\autoref{tab:long_corr}) 
of the parameters 
indicates significant parameter indeterminability. When the input
consists entirely of impulses, the best parameters are not one particular
set, but a spectrum. Note that the correlation is in parameters whose priors
are completely independent. It is possible that varying the type of input could
give improved results, although informal tests did not show significant difference.

@@

The following tests all use less data, on par with the average length of an FMRI
session.  For this low noise case, the eleven realizations are shown in 
\autoref{fig:LowNoiseRealization}.
The bias introduced into the signal by preprocessing 
could have some effect on the resulting fit; thus the preprocessed signal is compared
to the true BOLD signal in \autoref{fig:PreprocessedLowNoise}.
Overall, \autoref{fig:PreprocessedLowNoise} shows that the preprocessing was 
effective at removing trends. The spline did struggle to match the signal 
torward the end. This slight drift effect was caused by the final median being 
well above the base
level. The estimated BOLD signal is compared to the true BOLD signal in \autoref{fig:FitComparisonLowNoise}.
At several time points the particle filter successfully filtered the input. 
For instance, in the last 30 seconds
the estimates stay flat in spite of the preprocessed data drifting off. By
this point, the algorithm had converged sufficiently to prevent such inexplicable movement.
A similar circumstance occurs at 100 seconds in. A combination of 
noise and preprocessing biased the results toward a peak signal above the true peaks. 
The final parameter sets are shown in 
\autoref{tab:LowNoiseResults}.

@@

\begin{figure}
\subfigure[Converging histogram for $q$, $s$, $f$, and $BOLD$ of the first run, low noise simulation.]
{\includegraphics[clip=true,trim=7cm 3cm 6cm 3cm, width=16cm]{images/converge_lownoise3}}
\label{fig:LowNoiseHist}
\caption{Convergence of the first run from \autoref{tab:LowNoiseResults}. The bars represent
a histogram, where darker bars indicate more particles in that bin.}
\end{figure}

@@

The particles converged much faster when more noise was present (\autoref{fig:ConvergenceRuns1}, 
\autoref{fig:ConvergenceRuns2}). This also caused significantly more resampling which 
is the explanation for the percieved jumps in the histograms. 
The parameters arrived at for all 11 runs are shown in \autoref{tab:HighNoiseResults}.
Clearly the additional noise resulted in much more sporadic results. This is often the 
result when the particle filter converges too fast, in this case the result of the 
weighting function's variance being smaller than the measurement noise ($0.005$ vs. $0.01$). 
The error clearly suffers due to this effect (\autoref{tab:HighNoiseResults}).

>>>>

To test how well the particle filter would do with plenty of data, and 
determine the inherent variance in the model parameters, a very long simulation
with randomly generated impulse stimuli was created. The preprocessed timeseries
and the final estimate are shown in \autoref{fig:long_converge};
the final covariance matrix of the parameters is in \autoref{tab:long_cov}.
Note that even though the BOLD response converged well (\autoref{fig:long_converge}),
the parameters still have significant correlation (\autoref{tab:long_corr}). 
Based on the histograms in the first 500 seconds, the parameters converged to
their final values well before the end.
Although this is only a single test, the correlation (\autoref{tab:long_corr}) 
of the parameters 
indicates significant parameter indeterminability. When the input
consists entirely of impulses, the best parameters are not one particular
set, but a spectrum. Note that the correlation is in parameters whose priors
are completely independent. It is possible that varying the type of input could
give improved results, although informal tests did not show significant difference.
In spite of the noisy input (green line in \autoref{fig:long_converge}), the estimates
of the BOLD were actually very close to the true (noise-free) BOLD signal.

@@

\begin{figure}
\subfigure[Converging histogram for $q$, $s$, $f$, and $BOLD$ of the first run, low noise simulation.]
{\includegraphics[clip=true,trim=7cm 3cm 6cm 3cm, width=16cm]{images/converge_lownoise3}}
\caption{Convergence of the first run from \autoref{tab:LowNoiseResults}. The bars represent
a histogram, where darker bars indicate more particles in that bin.}
\label{fig:LowNoiseHist}
\end{figure}

The following tests all use less data (shorter sequences), on par with 
the average length of an FMRI session. For this low noise case 
$(\sigma_y = 0.001,\ \sigma_x = .0005)$, the eleven realizations are shown in 
\autoref{fig:LowNoiseRealization}.
The bias introduced into the signal by preprocessing 
could have some effect on the resulting fit; thus the preprocessed signal is compared
to the true BOLD signal in \autoref{fig:PreprocessedLowNoise}.
Overall, \autoref{fig:PreprocessedLowNoise} shows that the preprocessing was 
effective at removing trends, although the spline did struggle to match the signal 
torward the end. This slight drift effect was caused by the final median being 
well above the base
level. 
At several time points the particle filter successfully filtered the input
(\autoref{fig:FitComparisonLowNoise}).
For instance, in the last 30 seconds
the estimates stay flat in spite of the preprocessed data drifting off. By
this point, the algorithm had converged sufficiently to prevent such inexplicable movement.
A similar circumstance occurs at around 100 seconds in. A combination of 
noise and preprocessing biased the results toward a peak signal above the true peaks. 
The final parameter sets are shown in 
\autoref{tab:LowNoiseResults}.

@@

@@

The particles converged much faster when more noise was present (\autoref{fig:LowNoiseHist} vs. 
\autoref{fig:ConvergenceRuns1}, \autoref{fig:ConvergenceRuns2}). 
This also caused significantly more resampling which 
is the explanation for the percieved jumps in the histograms. 
The parameters arrived at for all 11 runs are shown in \autoref{tab:HighNoiseResults}.
Clearly the additional noise resulted in much more sporadic results. This is often the 
result when the particle filter converges too fast, in this case the result of the 
weighting function's variance being smaller than the measurement noise ($0.005$ vs. $0.01$). 
The error clearly suffers due to this effect (\autoref{tab:HighNoiseResults}).
